Leadership across cultures and contexts: an exploratory analysis

### 2020-05-07 leadership_across_cultures_contexts.Rmd
Polish edits in results section and first sections of discussion.

The GLOBE CEO Study defined six "global leadership behaviors" developed from 21 "primary leadership behaviors." Among the global leadership behaviors, they found that charismatic leadership behavior was consistently the most important behavioral trait regarding firm performance and dedication of top management team (TMT) members, followed by team-oriented behavior and humane-oriented behavior. Participative leadership was moderately related to TMT dedication, however was not related to firm performance. The most successful CEOs are those who are rated as more visionary, performance oriented, and decisive. Furthermore, results strongly suggest the fit between a CEO's behavioral profile and the expectations of leadership within a society strongly predict TMT dedication and firm performance. By developing higher-order features of leadership behavior and situating leadership functions within more detailed classification frameworks, insights from managerial studies can more effectively be coupled with approaches from acroos behavioral science.

<!-- ADDED CONCLUSION PARAGRAPH/STATEMENT.  -->
In contrasting the functions leaders serve in smaller-scale, non-WEIRD populations from larger-scale, WEIRD ones, community leaders in non-WEIRD settings are more likely to serve context-specific functions based on role-specialization [see @smith_leadership_2016]. The hunt leader does not necessarily have authority over military raids and the lineage head does not necessarily have authority over ritual ceremonies. In WEIRD settings however, community, organizational, and political leaders often have decision-making authority across multiple contexs. In the United States, despite a designed separation of power across branches of government, presidents, governors, and mayors oversee governmental bodies and have substantial influence in economics, social and legal policies, and international and inter-regional relations. Community leadership functions in WEIRDer populations will likely be increasingly focused on economic contexts, creatign and managing organizational structures, and inter-group relations, whereas in less WEIRD contexts community leaders will primarily function to resolve interpersonal conflicts and manage intra-group social dynamics.
<!-- Could add ethnographic examples for SSS and LSS  -->

@@

Evolutionary theoretical models and studies often seek to integrate findings within a broad framework of human cultural diversity. The universality of human leadership systems and the incredible diversity among the qualities of leaders across cultures and contexts and the functions leaders serve is difficult to capture. Scholars are then left to haphazardly self-select ethnographic cases or rely on the self-selected reviews of others for generalizations [e.g., @garfield_review_2019; @von_rueden_leadership_2015]. Here, we seek to provide the systematic and near comprehensive view of the qualities and functions of leaders across human societies from a representative cross-cultural sample of nonindustrial, non-WEIRD populations. We also provide the first comparative view describing the costs and benefits of leadership for both leaders and followers in an evolutionary biological framework. We also aim to explore how these dimensions of leadership vary within populations across social contexts and across populations with different subsistence strategies, as well as by leader sex and contintental region.

@@

Our third goal was to assess the extent to which each of the leadership trait and cost/benefit variables varied by four theoretically important factors for which we had data for all rows: subsistence strategy, continental region, group context, and leader sex.

@@

Our fifth goal was to assess variation in our "feature" dimensions by subsistence strategy, continental region, group context, and leader sex.

@@

The number of text records on leadership retrieved for a particular culture was weakly correlated with the total number of pages of ethnography in the eHRAF (r = `r signif(cor_pages_tr$estimate, 2)`, p = `r signif(cor_pages_tr$p.value, 2)`). A loess regression suggested that the number of text records initially increased with the total number of pages of ethnography, plateauing at about 35 records around 4500 pages (see Supplementary Figure \@ref(fig:eHRAFpages)).

We fit `r length(all_study_vars)` logistic mixed effects models with each binary study variable as the outcome; fixed effects for publication date, any female authors (True/False), and total pages of ethnography from each culture; and random intercepts for ethnographers nested within cultures. We then corrected for multiple comparisons by setting a false discovery rate of 5% [@benjamini1995]. After correction, none of the coefficients of female authorship or total pages of ethnography were significant, but the publication date coefficient was significant in nine models. More recent publications were more likely to report evidence for *Oratory skill*, *Group representative*, *Group determination/cohesiveness*, and *Knowledgeable/intelligent*, and less likely to report evidence for leaders imposing costs (*Punishment*, *Killer*) or receiving benefits (*Social services*, *Food*, and *Reduced risk of harm*). See Supplementary Figure \@ref(fig:pubdate) and supplementary information for details.

@@

The most common functions of leaders, documented in over 70% of cultures, were resolving conflicts, providing various social functions, organizing various cooperative activities, and providing counsel or direction in follower's affairs. Leaders were also commonly described as serving as a representatives in inter-group interactions, managing the distribution of resources, and providing context-specific leadership such as military command and in providing ritual services. In about half of the cultures sampled here leaders punished other group members and managed economic systems. Notable infrequent functions included controlling group movement and migrations.

The most common qualities of leaders were having high social status (documented in over 90% of cultures), being experienced or accomplished, and being described as knowledgeable or intelligent (documented in approximately 80% of cultures). Also, in about 60% of the cultures sampled here leaders were wealthy, nepotistic, possessed supernatural qualities, were generous, or maintained coercive authority over followers. Notable infrequent qualities of leaders in these data included physical formidability, having many social contacts, and being feared (documented in about 35% of cultures) and charisma and fairness (documented in about 20% of cultures).

@@

The frequency of leader and follower costs and benefits were estimated similarly to the leader qualities and functions described above. In these data leaders most commonly received benefits in the form of greater access to material resources (documented in about 70% of cultures) and increased social status (documented in about 65% of cultures). Mating benefits and receiving social services were also relatively frequent benefits of leadership. In general, the costs tracked the benefits. Leaders were often described as losing their high social status (documented in about 45% of cultures) and greater access to resources (documented in about 40% of cultures). See Figure \@ref(fig:varsupportcb).

Followers on the other hand most commonly benefited from leadership systems through receiving material resources (documented in about 50% of cultures) but otherwise most frequently through social services provided (documented in about 35% of cultures). Mating and fitness benefits were relatively rare for followers. Followers were most frequently described as incurring costs in the form of material resources lost (documented in about 40% of cultures) and in providing social services (documented in about 25% of cultures). See Figure \@ref(fig:varsupportcb).

@@

To investigate variation in our `r length(all_study_vars)` dimensions of leadership, we fit a multiple logistic mixed effects regression model of each dimension as a function of four factors: group structure, subsistence strategy, continental region, and leader sex, with random intercepts for author nested within culture. We then compared the AIC value of each model to a model with only the fixed and random intercepts. Models of leadship dimensions whose AIC values that showed less than a two-point reduction relative to the intercept-only model were deemed to not vary by these factors, whereas those whose AIC values were less two or more were deemed to vary.

@@

We first examined the `r commondim_n` leader dimensions for which evidence was found in at least 60% of cultures. There were `r universals_n` that did not show evidence of varying by the four factors. Because evidence for these dimensions was common across cultures, the `r universals_n` unvarying dimension are condidates for universal dimensions of leadership. See Table \@ref(tab:leaderuniversals).

@@

df_compare2 %>% 
  filter(AIC_diff >= -2 & Estimate > 0.6) %>% 
  mutate(
    Type = map_chr(var, ~ str_split(., '_')[[1]][[1]]),
    Type = str_replace(Type, '\\.', ' '),
    Type = str_to_title(Type),
    `Percent of cultures` = 100*Estimate
  ) %>% 
  select(Type, Variable, `Percent of cultures`) %>% 
  arrange(-`Percent of cultures`) %>% 
  kable(digits = 1, caption = "Candidate universal dimensions of leadership. These dimensions were common in the ethnographic record (appearing in at least 60% of cultures), and did not appear to vary by subsistence strategy, continental region, leader sex, or group context.")

@@

```{r heatmapdrop1, fig.width=8, fig.cap="Leadership dimensions that varied by subsistence strategy, continental region, leader sex, and/or group context. Colors represent the relative change in AIC value after dropping that factor from a multiple logistic mixed effects regression model. Yellow: factor improved fit relative to other factors. Black: factor did not improve fit relative to other factors. Delta AIC values were standardized by rows, and therefore can only be compared within rows."}
heatmap_drop1
```

For each of our four factors, we then investigated which levels of the factor was associated with higher or lower levels of evidence. For instance, for subsistence strategy, there were three leadership dimensions in which it played an important role: No coercive authority, Provide subsistence, and Distribute resources. See Figure \@ref(fig:heatmapdrop1). We plotted a heatmap that displayed the relative marginal means for each subsistence strategy for each of the three leadership dimensions (see Figure S\@ref(fig:heatmapsubsis)). Hunter-gatherers had particularly high levels of evidence for No coercive authority, relative to the other subsistence strategies (and averaging over the levels of the other three factors), and Pastorlists had relatively low evidence for Distribute resources. For Region, see Figure, S\@ref(fig:heatmapregion); for Leader sex, see Figure S\@ref(fig:heatmapsex); and for Group context, see Figure S\@ref(fig:heatmapgroups).

@@

*High status* was the most frequently identified leader quality in the sample, with `r statusPercTxt`% of text records and `r statusPercCult`% of cultures containing evidence that leaders had high social status. To explore factors that distinguished these records from those with no evidence for high status, we first created a document-term matrix (DTM) of all "informative" words in our corpus of texts and the frequency with which they occurred in each text record (see supplementary information for details). We then fit an elasticnet logistic regression model (with lasso penalty) of *High status* (0/1) as a function of the frequencies of all `r ncol(leader_dtm)-1` words. Words that were strong positive predictors epitomized the semantic content of the text records which provided evidence of high status as a quality of leaders. These included *`r rev(highstatus_plot$data$Word)[1]`* and *`r rev(highstatus_plot$data$Word)[2]`*, which were the strongest predictive terms; *`r rev(highstatus_plot$data$Word)[3]`* and *`r rev(highstatus_plot$data$Word)[4]`* were also relatively strong predictors; and *`r rev(highstatus_plot$data$Word)[5]`* and *`r rev(highstatus_plot$data$Word)[6]`*, which were also positive predictors. Notable terms which offered predictive value but with near-zero coefficients included, *influence*, *authority*, *headman*, and *chief*. See Figure S\@ref(fig:highstatusplot).

To determine which leader qualities and functions were associated with evidence for high status, we fit a logistic elasticnet model of *High status* as a function of all quality and function variables, with the lasso penalty ($\alpha = 1$). The non-zero coefficients indicated that *Punishment* was a negative predictor, all other functions, including *Organize cooperation*, *Ritual functions*, and *Provide counsel/direction* were non-predictors, and *Wealthy*, *Strategic nepotism*, *Age*, *Experienced/accomplished*, and several others, were positive predictors. See Figure \@ref(fig:elastichighstatus).

```{r elastichighstatus, fig.width=10, fig.cap="Coefficients of logistic elasticnet regression models [@friedman_regularization_2010] of evidence for *High Status*. A: Coefficients indicate the words whose frequencies in each text record best predicted evidence for *High status* in each text record. B: Non-zero coefficients of all quality and function variables that best predicted evidence for *High status*. Both models used the lasso penality ($\\alpha=1$), with $\\lambda_{1se}$ chosen by cross-validation (1 SE from \\lambda_{min}). Red values are positive predictors; purple values are negative predictors. Coefficents = 0 not displayed. X-axes are odds ratios on a log 10 scale."}

@@

We then investigated if the proportion of text records with evidence for *High status* differed among the different levels of group structure, subsistence strategies, continental regions, and leader sex (again using a logistic mixed effects model with these variables as fixed effects and authors nested within cultures as random intercepts). There were modest differences. Hunter-gatherers and pastoralists were more likely to provide evidence for high status leaders, as were South America and East Eurasia. Female leaders were also more likely to be identified as high status. Somewhat counter-intuitively, the proportion of text records indicating high status leaders was lowest at the highest level of political organization, perhaps because it would have been superfluous to note that kings and chiefs were high status. See Figure \@ref(fig:groupstatus). For the table of coefficients, see Supplementary Table X.

@@

@@

Here we report the cluster analysis of variables that used the correlation distance metric ($1 - cor$), and the Ward agglomeration algorithm (for results from the other methods, see the supplementary information). Figure \@ref(fig:cluster)A displays a dendrogram from the cluster analysis of the `r length(quality_vars)` leader quality variables and Figure \@ref(fig:cluster)B displays a dendrogram from the cluster analysis of the `r length(function_vars)` leader function variables. Each cluster analysis includes two estimates of significance for how strongly each cluster within the dendrogram is supported by the data. We rely on the AU (Approximately Unbiased) *p* values, which are computed by multiscale bootstrap resampling and represented as percentages (e.g., clusters with AU values $>95$ are strongly supported, and the top-level clusters are automatically highlighted by rectangles).

@@

We identify these named subclusters as "features" of leader qualities and functions in the ethnographic record. In naming these features, we gave extra weight to variables that were strongly supported in the sample (Figure \@ref(fig:varsupport)), and therefore not all variables within these clusters were clearly related to the themes of each feature.

We then fit separate logistic mixed effects models of each feature as a function of subsistence strategy, continental region, group context, and leader sex, with random effects for author nested within culture, comparing these to intercept-only models, and selecting all models whose AIC values were at least two points less than the intercept-only models. See Figures S\@ref(fig:heatmapfeaturesubsis), S\@ref(fig:heatmapfeatureregion), S\@ref(fig:heatmapfeaturesex), and S\@ref(fig:heatmapfeaturegroups).

@@

This exploratory study systematically evaluated evidence for `r length(all_study_vars)` dimensions of leadership in a stratified random sample of the ethnographic record of `r nrow(leader_cult)` largely non-industrial (non-WEIRD) cultures that included hunter-gatherers, pastorlists, horticulturalists, and agriculturalists. The quality and function dimensions were *not* determined prior to the study but were instead identified and coded during multiple passes through `r nrow(text_records)` texts on leadership obtained from the eHRAF (see Methods section). Results revealed universal dimensions of human leadership as well as important variation by group context, subsistence type, continental region, and sex of the leader.

We found strong evidence that several dimensions of leadership were universal across cultures, and across group contexts within cultures. We found evidence that leaders function to resolve conflict, organization cooperation, and provide counsel and direction in over 70% of cultures, and are seen as high status, knowledgeable and intelligence, and experienced and accomplished in about 80% or more of cultures. See Figure \@ref(fig:varsupport). Leaders benefitted materially, reproductively, and socially in 50% or more of cultures. See Figure \@ref(varsupportcb). Some common leader dimensions also did not vary by continental region, subsistence strategy, group context, or leader sex. These included qualities, such as being knowledgeable and intelligence, wealthly, strategically nepotistic, and generous, and functions, such as resolving conflict and representing the group. These dimensions are therefore candidates for universal dimensions of human leadership. By "universal", we do not mean that all leaders have all these traits. Instead, we mean that these traits are found in some leaders across all cultures and contexts.

Many leadership dimensions did vary across cultures and contexts, however, including functions such as organizing cooperation, providing counsel and direction and prosocial investment, and qualities such as high status and aggressiveness. See Figure \@ref(fig:heatmapdrop1)

Most discussions of leadership in the ethnographic record pertained to male leaders of community-level political groups, with leadership in residential subgroups and kin groups relatively more important in hunter-gatherer and pastoralist societies, and supracommunity political groups more important in horticultural and agricultural societies. Leadership in all social contexts, including economic, military, and religious, was present across all subsistence strategies, however. See Figure \@ref(fig:grouptype). The relative unimportance of supracommunity leadership in hunter-gatherers and pastorlists, and the relative importance of leadership in residential subgroups, highlights a major difference with WEIRD societies that more closely resemble the agricultural societies in our sample that show the opposite pattern.

@@

Across cultures and contexts, when ethnographic texts described female leaders exclusively, they were relatively more likely to be described as older, of high status, married to a high-quality spouse in a polygynous marriage, and the recipients of various mating benefits. This suite of characteristics is consistent with discussions of high status first-wives who gain and maintain increased social influence across the lifespan by leveraging their extended kin and social networks [@goodale_tiwi_1971; @brown_her_1985], and supports the view of positive assortative mating of high status leaders @garfield_investigating_2019 documented among elected leaders in a small-scale society with relative gender-egalitarianism. Evidence for male leaders (relative to evidence for female leaders or evidence that was not sex-specific) was indeed linked to group-level and political functions and male leaders were relativey more likely to provide social services, control political structure, and were at greater risk of losing their high social standing. Female leaders, relative to males, were also more likely to lack coercive authority, which is consistent with much of leadership studies suggesting dominant and aggressive women are less likely to be preferred as leaders, unlike dominent men who are sometimes prefered as leaders [@garfield_investigating_2019; @williams_subtle_2016]. See Figure S\@ref(fig:heatmapsex).

>>>>

The GLOBE CEO Study defined six "global leadership behaviors" developed from 21 "primary leadership behaviors." Among the global leadership behaviors, they found that charismatic leadership behavior was consistently the most important behavioral trait regarding firm performance and dedication of top management team (TMT) members, followed by team-oriented behavior and humane-oriented behavior. Participative leadership was moderately related to TMT dedication, however was not related to firm performance. The most successful CEOs are those who are rated as more visionary, performance oriented, and decisive. Furthermore, results strongly suggest the fit between a CEO's behavioral profile and the expectations of leadership within a society strongly predict TMT dedication and firm performance. By developing higher-order features of leadership behavior and situating leadership functions within more detailed classification frameworks, insights from managerial studies can more effectively be coupled with approaches from across behavioral science.

<!-- ADDED CONCLUSION PARAGRAPH/STATEMENT.  -->
In contrasting the functions leaders serve in smaller-scale, non-WEIRD populations from larger-scale, WEIRD ones, community leaders in non-WEIRD settings are more likely to serve context-specific functions based on role-specialization [see @smith_leadership_2016]. The hunt leader does not necessarily have authority over military raids and the lineage head does not necessarily have authority over ritual ceremonies. In WEIRD settings however, community, organizational, and political leaders often have decision-making authority across multiple contexts. In the United States, despite a designed separation of power across branches of government, presidents, governors, and mayors oversee governmental bodies and have substantial influence in economics, social and legal policies, and international and inter-regional relations. Community leadership functions in WEIRDer populations will likely be increasingly focused on economic contexts, creating and managing organizational structures, and inter-group relations, whereas in less WEIRD contexts community leaders will primarily function to resolve interpersonal conflicts and manage intra-group social dynamics.
<!-- Could add ethnographic examples for SSS and LSS  -->

@@

Evolutionary theoretical models and studies often seek to integrate findings within a broad framework of human cultural diversity. The universality of human leadership systems and the incredible diversity among the qualities of leaders across cultures and contexts and the functions leaders serve is difficult to capture. Scholars are then left to haphazardly self-select ethnographic cases or rely on the self-selected reviews of others for generalizations [e.g., @garfield_review_2019; @von_rueden_leadership_2015]. Here, we seek to provide the systematic and near comprehensive view of the qualities and functions of leaders across human societies from a representative cross-cultural sample of nonindustrial, non-WEIRD populations. We also provide the first comparative view describing the costs and benefits of leadership for both leaders and followers in an evolutionary biological framework. We also aim to explore how these dimensions of leadership vary within populations across social contexts and across populations with different subsistence strategies, as well as by leader sex and continental region.

@@

Our third goal was to assess the extent to which each of the researcher-coded dimensions of leadership varied by four theoretically important factors for which we had data for all rows: group context, subsistence strategy, continental region, and leader sex.

@@

Our fifth goal was to assess variation in our "feature" dimensions by group context, subsistence strategy, continental region, and leader sex.

@@

The number of text records on leadership retrieved for a particular culture was weakly correlated with the total number of pages of ethnography in the eHRAF (r = `r signif(cor_pages_tr$estimate, 2)`, p = `r signif(cor_pages_tr$p.value, 2)`). A loess regression suggested that the number of text records initially increased with the total number of pages of ethnography, plateauing at about 35 records around 4500 pages (see Figure S\@ref(fig:eHRAFpages)).

We fit `r length(all_study_vars)` logistic mixed effects models with each binary study variable as the outcome; fixed effects for publication date, any female authors (True/False), and total pages of ethnography from each culture; and random intercepts for ethnographers nested within cultures. We then corrected for multiple comparisons by setting a false discovery rate of 5% [@benjamini1995]. After correction, none of the coefficients of female authorship or total pages of ethnography were significant, but the publication date coefficient was significant in nine models. More recent publications were more likely to report evidence for *Oratory skill*, *Group representative*, *Group determination/cohesiveness*, and *Knowledgeable/intelligent*, and less likely to report evidence for leaders imposing costs (*Punishment*, *Killer*) or receiving benefits (*Social services*, *Food*, and *Reduced risk of harm*). See Figure S\@ref(fig:pubdate) and supplementary information for details.

@@

The most common functions of leaders, documented in over 70% of cultures, were resolving conflicts, providing various social functions, organizing various cooperative activities, and providing counsel or direction in follower's affairs. Leaders were also commonly described as serving as a representatives in inter-group interactions, managing the distribution of resources, and providing context-specific leadership such as military command and in providing ritual services. In about half of the cultures sampled here leaders punished other group members and managed economic systems. Notable infrequent functions included controlling group movement and immigration.

The most common qualities of leaders were having high social status (documented in over 90% of cultures), being experienced or accomplished, and being described as knowledgeable or intelligent (documented in approximately 80% of cultures). Also, in about 60% of cultures sampled leaders were wealthy, nepotistic, possessed supernatural qualities, were generous, or maintained coercive authority over followers. Notable infrequent qualities of leaders included physical formidability, having many social contacts, and being feared (documented in about 35% of cultures) and charisma and fairness (documented in about 20% of cultures).

@@

The frequency of leader and follower costs and benefits were estimated similarly to the leader qualities and functions described above. In these data leaders most commonly received benefits in the form of greater access to material resources (documented in about 70% of cultures) and increased social status (documented in about 65% of cultures). Mating benefits and social services benefits were also relatively frequent benefits of leadership. In general, the costs tracked the benefits. Leaders were often described as losing their high social status (documented in about 45% of cultures) and greater access to resources (documented in about 40% of cultures). See Figure \@ref(fig:varsupportcb).

Followers on the other hand most commonly benefited from leadership systems through receiving material resources (documented in about 50% of cultures) but otherwise most frequently through receiving social services (documented in about 35% of cultures). Mating and fitness benefits were relatively rare for followers. Followers were most frequently described as incurring costs in the form of material resources lost (documented in about 40% of cultures) and in providing social services (documented in about 25% of cultures). See Figure \@ref(fig:varsupportcb).

@@

To investigate variation in our `r length(all_study_vars)` dimensions of leadership, we fit a multiple logistic mixed effects regression model of each dimension as a function of four factors: group context, subsistence strategy, continental region, and leader sex, with random intercepts for author nested within culture. We then compared the AIC value of each model to a model with only the fixed and random intercepts. Models of leadership dimensions whose AIC values that showed less than a two-point reduction relative to the intercept-only model were deemed to not vary by these factors, whereas those whose AIC values were less two or more were deemed to vary.

@@

We first examined the `r commondim_n` leadership dimensions for which evidence was found in at least 60% of cultures. There were `r universals_n` that did not show evidence of varying by the four factors. Because evidence for these dimensions was common across cultures, the `r universals_n` unvarying dimension are candidates for universal dimensions of human leadership. See Table \@ref(tab:leaderuniversals).

@@

df_compare2 %>% 
  filter(AIC_diff >= -2 & Estimate > 0.6) %>% 
  mutate(
    Type = map_chr(var, ~ str_split(., '_')[[1]][[1]]),
    Type = str_replace(Type, '\\.', ' '),
    Type = str_to_title(Type),
    `Percent of cultures` = 100*Estimate
  ) %>% 
  select(Type, Variable, `Percent of cultures`) %>% 
  arrange(-`Percent of cultures`) %>% 
  kable(digits = 1, caption = "Candidate universal dimensions of leadership. These dimensions were common in the ethnographic record (appearing in at least 60% of cultures), and did not appear to vary by group context, subsistence strategy, continental region, or leader sex.")

@@

```{r heatmapdrop1, fig.width=8, fig.cap="Leadership dimensions that varied by group context, subsistence strategy, continental region, and/or leader sex. Colors represent the relative change in AIC value after dropping that factor from a multiple logistic mixed effects regression model. Yellow: factor improved fit relative to other factors. Black: factor did not improve fit relative to other factors. Delta AIC values were standardized by rows, and therefore can only be compared within rows."}
heatmap_drop1
```

For each of our four factors, we then investigated which levels of the factor was associated with higher or lower levels of evidence for the leadership dimension. For instance, for subsistence strategy, there were three leadership dimensions in which it played an important role: *No coercive authority*, *Provide subsistence*, and *Distribute resources* (See Figure \@ref(fig:heatmapdrop1)). We plotted a heatmap that displayed the relative marginal means for each subsistence strategy for each of the three leadership dimensions (see Figure S\@ref(fig:heatmapsubsis)). Hunter-gatherers had particularly high levels of evidence for *No coercive authority*, relative to the other subsistence strategies (and averaging over the levels of the other three factors), and Pastoralists had relatively low evidence for *Distribute resources*. For Group context, see Figure S\@ref(fig:heatmapgroups); for Region, see Figure, S\@ref(fig:heatmapregion); and for Leader sex, see Figure S\@ref(fig:heatmapsex).

@@

*High status* was the most frequently identified leader quality in the sample, with `r statusPercTxt`% of text records and `r statusPercCult`% of cultures containing evidence that leaders had high social status. To explore factors that distinguished these records from those with no evidence for high status, we first created a document-term matrix (DTM) of all "informative" words in our corpus of texts and the frequency with which they occurred in each text record (see supplementary information for details). We then fit an elasticnet logistic regression model (with the lasso penalty, $\alpha = 1$) of *High status* (0/1) as a function of the frequencies of all `r ncol(leader_dtm)-1` words. Words that were strong positive predictors epitomized the semantic content of the text records which provided evidence of high status as a quality of leaders. These included *`r rev(highstatus_plot$data$Word)[1]`* and *`r rev(highstatus_plot$data$Word)[2]`*, which were the strongest predictive terms; *`r rev(highstatus_plot$data$Word)[3]`* and *`r rev(highstatus_plot$data$Word)[4]`* were also relatively strong predictors; and *`r rev(highstatus_plot$data$Word)[5]`* and *`r rev(highstatus_plot$data$Word)[6]`*, which were also positive predictors. Notable non-predictor terms included, *influence*, *authority*, *headman*, and *chief*. See Figure \@ref(fig:elastichighstatus)A.

To determine which leader qualities and functions were associated with evidence for high status leaders, we fit an elasticnet logistic model of *High status* as a function of all quality and function variables, (with the lasso penalty, $\alpha = 1$). The non-zero coefficients indicated that *Punishment*, the only identified function, was a negative predictor, whereas
<!-- all other functions, including *Organize cooperation*, *Ritual functions*, and *Provide counsel/direction* were non-predictors, and  -->
*Wealthy*, *Strategic nepotism*, *Age*, *Experienced/accomplished*, and several other qualities, were positive predictors. See Figure \@ref(fig:elastichighstatus)B.

```{r elastichighstatus, fig.width=10, fig.cap="Coefficients of logistic elasticnet regression models [@friedman_regularization_2010] of evidence for *High Status*. A: Coefficients indicate the words whose frequencies in each text record best predicted evidence for *High status* in each text record. B: Non-zero coefficients of all quality and function variables that best predicted evidence for *High status*. Both models used the lasso penality ($\\alpha=1$), with $\\lambda_{1se}$ chosen by cross-validation (1 SE from $\\lambda_{min}$). Red values are positive predictors; purple values are negative predictors. Coefficents = 0 not displayed. X-axes are odds ratios on a log 10 scale."}

@@

We then investigated if the proportion of text records with evidence for *High status* differed among the different levels of group context, subsistence strategies, continental regions, and leader sex (again using a logistic mixed effects model with these variables as fixed effects and authors nested within cultures as random intercepts). There were modest differences. Hunter-gatherers and pastoralists were more likely to provide evidence for high status leaders, as were South America and East Eurasia. Female leaders were also more likely to be identified as high status. Somewhat counter-intuitively, the proportion of text records indicating high status leaders was lowest at the highest level of political organization, perhaps because it would have been superfluous to note that kings and chiefs were high status. See Figure \@ref(fig:groupstatus). For the table of coefficients, see Supplementary Table X.

@@

*Coercive authority* was documented in about 57% of cultures. Unlike other more frequent leader qualities, which were generally positive or prosocial traits, *Coercive authority* was the most frequent negative or dominance-style leaders quality. We therefore chose to perform additional exploratory analyses with *Coercive authority* as the outcome in logistic elasticnet regression models similarly as described above for *High status* (see Figure \@ref(fig:elasticcoercive) caption for details).

Words that were strong predictors of leaders with coercive authority included, *power* and *authority* (unsurprisingly), as well as *territory*, *chief*, and *control*; the word *leadership* was the only strong negative predictor. See Figure \@ref(fig:elasticcoercive)A. Using all other quality and function variables as predictors identified *Punishment* as the strongest predictor of evidence for *Coercive authority*. The qualities of *Aggressiveness* and *Feared* were also positive predictors, as were the functions, *Political appointments*, *Control economics*, and *Control immigration*.

@@

Here we report the cluster analysis of variables that used the correlation distance metric ($1 - cor$), and the Ward agglomeration algorithm (for results from other methods, see the supplementary information). Figure \@ref(fig:cluster)A displays a dendrogram from the cluster analysis of the `r length(quality_vars)` leader quality variables and Figure \@ref(fig:cluster)B displays a dendrogram from the cluster analysis of the `r length(function_vars)` leader function variables. Each cluster analysis includes two estimates of significance for how strongly each cluster within the dendrogram is supported by the data. We rely on the AU (Approximately Unbiased) *p* values, which are computed by multiscale bootstrap resampling and represented as percentages (e.g., clusters with AU values $>95$ are strongly supported, and the top-level clusters are automatically highlighted by rectangles).

@@

We identify these named subclusters as "features" of leader qualities and functions in the ethnographic record. In naming these features, we gave extra weight to variables that were strongly supported in the sample (Figure \@ref(fig:varsupport)), and therefore not all variables within these clusters were clearly related to the themes of each feature. There were modest or no correlation between features, after aggregating leadership dimensions within features (see Figure \@ref(fig:featurecor)).

We then fit separate logistic mixed effects models of each feature as a function of group context, subsistence strategy, continental region, and leader sex, with random effects for author nested within culture, comparing these to intercept-only models, and selecting all models whose AIC values were at least two points less than the intercept-only models. See Figures S\@ref(fig:heatmapfeaturesubsis), S\@ref(fig:heatmapfeatureregion), S\@ref(fig:heatmapfeaturesex), and S\@ref(fig:heatmapfeaturegroups).

@@

This exploratory study systematically evaluated evidence for `r length(all_study_vars)` dimensions of leadership in a stratified random sample of the ethnographic record of `r nrow(leader_cult)` largely non-industrial (non-WEIRD) cultures that included hunter-gatherers, pastoralists, horticulturalists, and agriculturalists. The quality and function dimensions were *not* determined prior to the study but were instead identified and coded during multiple passes through `r nrow(text_records)` texts on leadership obtained from the eHRAF (see Methods section). Results revealed universal dimensions of human leadership as well as important variation by group context, subsistence type, continental region, and sex of the leader.

We found strong evidence that several dimensions of leadership were universal across cultures, and across group contexts within cultures. We found evidence that leaders function to resolve conflict, organization cooperation, and provide counsel and direction in over 70% of cultures, and are seen as high status, knowledgeable and intelligence, and experienced and accomplished in about 80% or more of cultures (see Figure \@ref(fig:varsupport)). Leaders benefited materially, reproductivly, and socially in 50% or more of cultures (see Figure \@ref(fig:varsupportcb)). Some common leadership dimensions also did not vary by group context, continental region, subsistence strategy, or leader sex. These included qualities, such as being knowledgeable and intelligence, wealthy, strategically nepotistic, and generous, and functions, such as resolving conflict and representing the group. These dimensions are therefore candidates for universal dimensions of human leadership. By "universal", we do not mean that all leaders have all these traits. Instead, we mean that these traits are found in some leaders across all cultures and contexts.

Many leadership dimensions did vary across cultures and contexts, however, including functions such as organizing cooperation, providing counsel and direction, and prosocial investment, and qualities such as high status and aggressiveness (see Figure \@ref(fig:heatmapdrop1)).

Most discussions of leadership in the ethnographic record pertained to male leaders of community-level political groups, with leadership in residential subgroups and kin groups relatively more common in hunter-gatherer and pastoralist societies, and supracommunity political groups more common in horticultural and agricultural societies. Leadership in all social contexts, including economic, military, and religious, was present across all subsistence strategies, however (see Figure \@ref(fig:grouptype)). The relative unimportance of supracommunity leadership in hunter-gatherers and pastoralists, and the relative importance of leadership in residential subgroups, highlights a trend within non-WEIRD populations in sociopolitical organizations, as these generally more sedentary populations with more intensified subsistence strategies more closely resemble sociopolitical organization of WEIRD societies. Thus, these results provide a fine-grained view of cultural transitions in leadership systems.

@@

Across cultures and contexts, when ethnographic texts described female leaders exclusively, they were relatively more likely to be described as older, of high status, married to a high-quality spouse in a polygynous marriage, and the recipients of various mating benefits. This suite of characteristics is consistent with discussions of high status first-wives who gain and maintain increased social influence across the lifespan by leveraging their extended kin and social networks [@goodale_tiwi_1971; @brown_her_1985], and supports the view of positive assortative mating of high status leaders @garfield_investigating_2019 documented among elected leaders in a small-scale society with relative gender-egalitarianism. Evidence for male leaders (relative to evidence for female leaders or evidence that was not sex-specific) was indeed linked to group-level and political functions and male leaders were relatively more likely to provide social services, control political structure, and were at greater risk of losing their high social standing. Female leaders, relative to males, were also more likely to lack coercive authority, which is consistent with much of leadership studies suggesting dominant and aggressive women are less likely to be preferred as leaders, unlike dominant men who are sometimes preferred as leaders [@garfield_investigating_2019; @williams_subtle_2016]. See Figure S\@ref(fig:heatmapsex).
